---
title: "Translatome F-M-ratio"
author: "Pascal Lafrenz"
date: "`r Sys.Date()`"
output: html_document
---

```{r}
library(ggpattern)
library(ggplot2)
library(ggpubr)
library(tidyverse)

out_dir <- "./translatome_analysis/"

if (!file.exists(out_dir)) {
  dir.create(out_dir, recursive = TRUE)
} 
```

# Loading data

Loading RPKM and metadata of male chicken head (fetal) and brain (adult) samples.

```{r}

chicken_female_male <- as.data.frame(read_csv("fwt_mwt_fpkm_chicken.csv"))
rownames(chicken_female_male) <- chicken_female_male$gene_id

sample_meta_fwt_mwt <- readr::read_delim("sample_meta_fwt_mwt.txt",delim = "\t")

```

## Targets

Loading metadata with mir-2954 targets.

```{r}
chick_gtf <- read_csv("FPKM_1_Z_linked_Upregulated_genes.csv")

chick_gtf <- chick_gtf %>%
  mutate(
    chicken_target = factor(chicken_target),
    mir_target_group = factor(paste0(chr_z_or_auto,"_", chicken_target))
  )
```

## Preprocessing

Transform the data into a useful format.

```{r}
chicken_female_male <- as.data.frame(chicken_female_male)
chicken_female_male$gene_id <- row.names(chicken_female_male)
chicken_female_male_longer <- pivot_longer(chicken_female_male,
                                    cols = !gene_id, 
                                    names_to = "library", 
                                    values_to = "fpkm")

update_res_table_chicken <- function(res_table, chick_gtf) {
  res_table <- res_table %>%
    as.data.frame() %>% 
    left_join(dplyr::select(chick_gtf, gene_id,expr_targets,chr_z_or_auto, mir_target_group, gene_biotype), by = "gene_id")
  return(res_table)
}

chicken_female_male_longer <- update_res_table_chicken(chicken_female_male_longer, chick_gtf)

chicken_female_male_longer <- left_join(chicken_female_male_longer,sample_meta_fwt_mwt,by=c("library"="Sample")) %>%
  mutate(Organ = case_when(
    Organ == "brain" ~ "brain adult",
    Organ == "head"  ~ "head fetal",
    TRUE ~ Organ  # Optional: keep the original value for other cases
  ))

chicken_female_male_longer <- chicken_female_male_longer %>%
  filter(fpkm > 0) %>%
  mutate(logFPKM = log2(fpkm))

chicken_female_male_longer
```

## Calculating ratios

Calculate the female-to-male ratios.

```{r}

FM_ratio <- chicken_female_male_longer %>% 
  group_by(Organ, gene_id, Type, gene_biotype, expr_targets, chr_z_or_auto) %>%
  dplyr::summarise(median_FPKM_female = median(logFPKM[Sex == "female"], na.rm = T),
                   median_FPKM_male = median(logFPKM[Sex == "male"], na.rm = T)) %>%
  filter(
    gene_biotype == "protein_coding",
      median_FPKM_female >= 1,
      median_FPKM_male >= 1
  ) %>%
  group_by(Type, Organ) %>%
  mutate(female_to_male_ratio = median_FPKM_female - median_FPKM_male,
         female_to_male_ratio_normalized = female_to_male_ratio - median(female_to_male_ratio[chr_z_or_auto == "Autosomal"], na.rm = T),
         expr_targets = factor(expr_targets, levels = c("not_exp_target", "exp_target"))
  ) %>%
  filter(chr_z_or_auto == "Z-linked")
FM_ratio
```
## Calculating statistics

To compare the ratios between target groups and data types, various statistics are calculated.

```{r}

FM_ratio_type_wilcox <- FM_ratio %>%
  group_by(Organ,expr_targets) %>%
  wilcox_test(female_to_male_ratio_normalized~Type,
              alternative = "two.sided", p.adjust.method = "BH") %>%
  mutate(y=1.65)
FM_ratio_type_wilcox

FM_ratio_target_wilcox <- FM_ratio %>%
  group_by(Organ,Type) %>%
  wilcox_test(female_to_male_ratio_normalized~expr_targets,
              alternative = "two.sided", p.adjust.method = "BH") %>%
  mutate(y=c(-2.25,-2.75,-2.25,-2.75),
         expr_targets=c("not_exp_target","exp_target","not_exp_target","exp_target"))
FM_ratio_target_wilcox

```

## Plotting

Here the distributions and the respective statistical comparisons are plotted.

```{r}

format_p_value <- function(p_value) {
  p_value_scientific <- formatC(p_value, format = "e", digits = 2)  # Keep 2 decimal places
  mantissa <- sub("e.*$", "", p_value_scientific)  # Extract mantissa
  exponent <- sub("^.*e", "", p_value_scientific)  # Extract exponent
  exponent <- as.integer(exponent)
  
  # Return formatted string for ggplot2 (expression-compatible)
  return(bquote(.(mantissa) ~ "×" ~ 10^.(exponent)))
}


ggplot(FM_ratio, aes(x = expr_targets, y = female_to_male_ratio_normalized, fill = expr_targets, pattern = Type)) +
  geom_boxplot_pattern(
    notch = TRUE, 
    outlier.shape = NA, 
    pattern_fill = "black", 
    pattern_density = 0.1,  # Adjust crosshatch density
    pattern_spacing = 0.02, # Adjust spacing of pattern
    pattern_linetype = 1
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "", x = "", y = "log2(F-M ratio)") +
  geom_segment(aes(x = 0.7, xend = 1.3, y = 1.5, yend = 1.5), color = "black", size = 0.5) +
  geom_segment(aes(x = 1.7, xend = 2.3, y = 1.5, yend = 1.5), color = "black", size = 0.5) +
  geom_segment(aes(x = 0.7, xend = 1.8, y = -2.4, yend = -2.4), color = "black", size = 0.5) +
  geom_segment(aes(x = 1.2, xend = 2.3, y = -2.9, yend = -2.9), color = "black", size = 0.5) +
  geom_text(data = FM_ratio_type_wilcox,
            aes(x = c(1, 2, 1, 2), y = y, label = sapply(p, format_p_value)),
            parse = TRUE, size = 5,
            inherit.aes = FALSE) +
  geom_text(data = FM_ratio_target_wilcox, aes(x = c(1.25, 1.75, 1.25, 1.75), y = y, label = sapply(p,format_p_value)), parse = T, size = 5) +
  scale_pattern_manual(name = "", values = c("ribo" = "stripe", "rna" = "none"),labels = c("ribo" = "RIBO", "rna" = "RNA")) +
  scale_fill_manual(name = "",
    values = c("not_exp_target" = "#DCE9F8", "exp_target" = "#6664AB"),
    labels = c("not_exp_target" = "Other-Z", "exp_target" = "Exp-Z")) +
  facet_grid(. ~ Organ) +
  geom_hline(yintercept = 0, color = "black", linetype = "dashed") +
  scale_x_discrete(labels = c("not_exp_target" = "Other-Z", "exp_target" = "Exp-Z")) +
  guides(fill = guide_legend(override.aes = list(pattern = "none"))) +
  theme_classic() +
  theme(text = element_text(size=16))


ggsave(paste0(out_dir,"/FM_ratio_norm.pdf"), dpi="retina",height = 9, width = 12)

```

## Significance from -1,0,1

Statistically significant deviations of the medians of these ratios from key reference values (e.g., 0.5 [log2 ratio of −1]; 1 [log2 ratio of 0]; and 2 [log2 ratio of 1]) were assessed using one-sample Wilcoxon signed rank tests. P-values were corrected for multiple testing using the Bonferroni procedure, with adjusted P < 0.05 indicating significance.

```{r}

# Tests of statistical significance for all 3 reference levels

## Reference:0
ratio_ref_test_0 <- FM_ratio %>%
  group_by(Organ, expr_targets, Type) %>% 
  wilcox_test(female_to_male_ratio_normalized ~ 1, mu = 0, alternative = "two.sided") %>%
  mutate(p.adj=round(p.adjust(p, method="BH"), 2))
## Reference:-1
ratio_ref_test_n1 <- FM_ratio %>%
  group_by(Organ, expr_targets, Type) %>% 
  wilcox_test(female_to_male_ratio_normalized ~ 1, mu = -1, alternative = "two.sided")%>%
  mutate(p.adj=round(p.adjust(p, method="BH"), 2))
## Reference:1
ratio_ref_test_p1 <- FM_ratio %>%
  group_by(Organ, expr_targets, Type) %>% 
  wilcox_test(female_to_male_ratio_normalized ~ 1, mu = 1, alternative = "two.sided")%>%
  mutate(p.adj=round(p.adjust(p, method="BH"), 2))

# Final plots
all_tests <- data.frame(tissue=c(ratio_ref_test_0$Organ, ratio_ref_test_n1$Organ, ratio_ref_test_p1$Organ),
                        expr_targets=c(ratio_ref_test_0$expr_targets, ratio_ref_test_n1$expr_targets, ratio_ref_test_p1$expr_targets),
                        ref=c(rep(0, nrow(ratio_ref_test_0)), rep(-1, nrow(ratio_ref_test_n1)), rep(1, nrow(ratio_ref_test_p1))), p=c(ratio_ref_test_0$p, ratio_ref_test_n1$p, ratio_ref_test_p1$p),
                        data_type=c(ratio_ref_test_0$Type,ratio_ref_test_n1$Type,ratio_ref_test_p1$Type),
                        p.adj=c(ratio_ref_test_0$p.adj, ratio_ref_test_n1$p.adj, ratio_ref_test_p1$p.adj))

all_tests$exp_targets <- factor(all_tests$expr_targets, levels=c("not_exp_target", "exp_target"))

## P-value heatmap
ggplot(all_tests, aes(x=data_type, y=ref, fill= p.adj<0.05)) +
  geom_tile(width = 0.5, height = 0.5) + 
  facet_wrap(.~ tissue+exp_targets, nrow = 1) + 
  theme_classic() + 
  scale_fill_manual(values = c("red", "lightgrey"),
                    breaks = c("FALSE", "TRUE"),
                    labels = c("> 0.05", "< 0.05")) +
  #geom_text(aes(y=ref,label=p.adj))+ 
  #theme(legend.position = "none") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1) ,
    panel.background = element_rect(fill = "white", color = NA),  # Set the panel background to white
    plot.background = element_rect(fill = "white", color = NA),   # Set the overall plot background to white
    text = element_text(size=16)
  ) +
  ylab("Reference log2(F-M-ratios)") +
  scale_x_discrete(labels = c("not_exp_target" = "Other-Z", "exp_target" = "Exp-Z")) +
  labs(title = "", x="", fill="Adj. P-val")
  
ggsave(paste0(out_dir,"/FM_ratio_norm_sig_expr.pdf"), dpi="retina",height = 9, width = 12)

```




