---
title: "Translatome F-M-TE-ratio"
author: "Pascal Lafrenz"
date: "`r Sys.Date()`"
output: html_document
---

```{r}
library(ggpattern)
library(ggplot2)
library(ggpubr)
library(tidyverse)

out_dir <- "./translatome_analysis/"

if (!file.exists(out_dir)) {
  dir.create(out_dir, recursive = TRUE)
} 
```


# Loading data

Loading RPKM and metadata of male chicken head (fetal) and brain (adult) samples.

```{r}

chicken_te <- as.data.frame(read_csv("te_fpkm_chicken.csv"))
rownames(chicken_te) <- chicken_te$gene_id

sample_meta_te <- readr::read_delim("sample_meta_te.txt",delim = "\t")

```


## Targets

Loading metadata with mir-2954 targets.

```{r}
chick_gtf <- read_csv("FPKM_1_Z_linked_Upregulated_genes.csv")

chick_gtf <- chick_gtf %>%
  mutate(
    chicken_target = factor(chicken_target),
    mir_target_group = factor(paste0(chr_z_or_auto,"_", chicken_target))
  )
```


## Preprocessing

Transform the data into a useful format.

```{r}
chicken_te <- as.data.frame(chicken_te)
chicken_te$gene_id <- row.names(chicken_te)
chicken_te_longer <- pivot_longer(chicken_te,
                                  cols = !gene_id, 
                                  names_to = "library", 
                                  values_to = "fpkm")

update_res_table_chicken <- function(res_table, chick_gtf) {
  res_table <- res_table %>%
    as.data.frame() %>% 
    left_join(dplyr::select(chick_gtf, gene_id,expr_targets,chr_z_or_auto, mir_target_group, gene_biotype), by = "gene_id")
  return(res_table)
}

chicken_te_longer <- update_res_table_chicken(chicken_te_longer, chick_gtf)

chicken_te_longer <- left_join(chicken_te_longer,sample_meta_te,by=c("library"="Sample")) %>%
  mutate(Organ = case_when(
    Organ == "brain" ~ "brain adult",
    Organ == "head"  ~ "head fetal",
    TRUE ~ Organ  # Optional: keep the original value for other cases
  ))

chicken_te_longer <- chicken_te_longer %>%
  filter(fpkm > 0) %>%
  mutate(logFPKM = log2(fpkm))

chicken_te_longer
```

## Calculating ratios

Calculate the female-to-male ratios of translational efficiencies ($$TE = \frac{RPF_{FPKM}}{RNA_{FPKM}}$$).

```{r}

te_ratio <- chicken_te_longer %>% 
  group_by(Organ, gene_id, Sex, gene_biotype, expr_targets, chr_z_or_auto) %>%
  dplyr::summarise(median_FPKM_ribo = median(logFPKM[Type == "ribo"], na.rm = T),
                   median_FPKM_rna = median(logFPKM[Type == "rna"], na.rm = T)) %>%
  filter(
    gene_biotype == "protein_coding",
      median_FPKM_ribo >= 1,
      median_FPKM_rna >= 1
  ) %>%
  group_by(Sex, Organ) %>%
  mutate(logTE = median_FPKM_ribo - median_FPKM_rna,
         logTE_norm = logTE - median(logTE[chr_z_or_auto == "Autosomal"], na.rm = T),
         expr_targets = factor(expr_targets, levels = c("not_exp_target", "exp_target"))
  ) %>%
  group_by(Organ, gene_id, gene_biotype, expr_targets, chr_z_or_auto) %>%
  dplyr::summarise(logTE_ratio = logTE[Sex == "female"] - logTE[Sex == "male"]) %>%
  group_by(Organ) %>%
  mutate(logTE_ratio_norm = logTE_ratio - median(logTE_ratio[chr_z_or_auto == "Autosomal"], na.rm = T))
  
te_ratio

```

## Calculating statistics

To compare the ratios between target groups, various statistics are calculated.

```{r}

te_ratio_target_wilcox <- rbind(
  te_ratio %>%
    filter(chr_z_or_auto == "Z-linked") %>%
    group_by(Organ) %>%
    wilcox_test(logTE_ratio_norm~expr_targets,
                alternative = "two.sided", p.adjust.method = "BH") %>%
    mutate(y=c(2.25,2.25),
           expr_targets=c("not_exp_target","exp_target")),
  te_ratio %>%
    filter(chr_z_or_auto == "Z-linked" & expr_targets == "exp_target" | chr_z_or_auto == "Autosomal") %>%
    group_by(Organ) %>%
    wilcox_test(logTE_ratio_norm~chr_z_or_auto,
                alternative = "two.sided", p.adjust.method = "BH") %>%
    mutate(y=c(2.75,2.75),
           expr_targets=c("not_exp_target","exp_target"))
)
te_ratio_target_wilcox
```

## Plotting

Here the distributions and the respective statistical comparisons are plotted.

```{r}

format_p_value <- function(p_value) {
  p_value_scientific <- formatC(p_value, format = "e", digits = 2)  # Keep 2 decimal places
  mantissa <- sub("e.*$", "", p_value_scientific)  # Extract mantissa
  exponent <- sub("^.*e", "", p_value_scientific)  # Extract exponent
  exponent <- as.integer(exponent)
  
  # Return formatted string for ggplot2 (expression-compatible)
  return(bquote(.(mantissa) ~ "×" ~ 10^.(exponent)))
}


ggplot(te_ratio, aes(x = chr_z_or_auto, y = logTE_ratio_norm, fill=expr_targets)) +
  geom_boxplot(notch = T, outlier.shape = NA) +
  labs(title = "", x = "", y = "Female-male-ratio log2(TE)") +
  geom_segment(aes(x = 0.7, xend = 2.3, y = 2.5, yend = 2.5), color = "black", size = 0.5) +
  geom_segment(aes(x = 1.7, xend = 2.3, y = 2, yend = 2), color = "black", size = 0.5) +
  geom_text(data = te_ratio_target_wilcox, aes(x = c(2,2,1.5,1.5), y = y, label = sapply(p,format_p_value)), parse = T, size = 5,inherit.aes = F) +
  geom_hline(yintercept = 0, color = "black", linetype = "dashed") +
  facet_grid(.~Organ) +
  scale_fill_manual(name = "",
    values = c("not_exp_target" = "#DCE9F8", "exp_target" = "#6664AB"),
    labels = c("not_exp_target" = "Other", "exp_target" = "Exp-Z")) +
  theme_classic() +
  theme(text = element_text(size=16))

ggsave(paste0(out_dir,"/te_chicken_norm.pdf"), dpi="retina",height = 9, width = 12)

```



